Bayesian Inference Using Sequential Monte-Carlo Algorithm for Dynamic System Models

### 2020-08-20 implementation/implementation.tex
Proofread

The inference framework code for the project was developed and tested in a local environment (macOS 10.15.6, see Table \ref{table:local_macine}) at first using a small input, then the functional version was deployed on compute node of ARCHER and Cirrus for parameter inference experiments and model selection tasks. The scaling-up experiment was performed on Cirrus. The profiling and its analysis were performed on a local machine. Some necessary development and debug were performed on the remote machines to ensure that experiments run correctly on the compute node.

The ODE model-related code and utilities were first developed to enable reusable functions, variables and data calls. It included the code for the ODE equations, ODE solver, data structures and data format transform functions, etc. By doing this in separate code files, we could change implementation options and activate ABC SMC easily in another `trigger' file, without creating definition and duplicated code for the models, observed data and parameter priors.

Some missing features were found when using \verb|pyABC|; therefore some modifications were made to the source code of \verb|pyABC|.

For example, the laboratory-measured data were not considered `complete' or `tidy' for modelling and analysis purpose: the available time points for cells and cytokines were different, e.g. cell number was not measured at 0.25 hpl and cytokines' expression was not measured at time point 120 hpl. These missing values were treated as `ignored' when calculating the distance between observed data and simulated data, and represented by \verb|NaN|s. However, in some distance calculations, \verb|NaN| was not acceptable. To cope with this, the built-in distance function was modified. Other modified code included some visualisation and result representation enhancements.

Analysis of the program output was also wrapped up into functions which can be easily called after an ABC SMC run is finished, which includes multiple visualisations and summary statistic of the results. For different tasks, there are also separated analysis code file. Other analysis tools used in the project include built-in database visualisation tool, R code and Microsoft Excel.

@@

    The ABC SMC inference can have widely different performance when using different hyperparameters and implementation options. To firstly test the ability of inference and observe the results of different options, synthetic trajectories with known true parameter values were used as target data. By doing this, we could directly compare the inferred parameters to the true values and compare the simulated trajectories to the true trajectories, then observe the result to see whether the inference is successful and how well the model fit the data. Also, by setting the same final threshold value, the efficiency under different options can be compared and help us to choose proper options when applying ABC SMC.

@@

    \caption[Synthetic data generated with known parameter values]%
    {Synthetic data generated with known parameter values (line plot), compared to experimental data (scatter plot). Known parameter values are obtained from a least square fitting of the observed data}
    \label{fig:infer_back_data}

@@

The perturbation kernel is called in the sampling of every particle and determines how the new perturbed particle is chosen; thus it is influential to both computation complexity and the resulting posterior distribution. Generally, a local perturbation kernel may face the risk of being stuck in local modes (e,g., local optimal), but it may need less computational operations or could generate a population with a higher acceptance rate if the successive epsilon values are close. A kernel with a wide variance, or spreading out in a large space could help in resolving the local optima problem by a more thorough exploration of the parameter space. However, it can be more computation-intensive and result in a lower acceptance rate. A desired optimal kernel should balance the trade-offs; their properties and criteria were discussed in \cite{ref:kernel}.

There are several common choices of perturbation kernels. Among them, the multivariate normal kernel and local M-nearest neighbour model is preferred to be applied to our models. A covariance matrix $\Sigma_t$ of accepted particles is calculated from the previous generation and used in multivariate normal kernel: $K(\theta|\theta_{t-1})\sim\mathcal{N}(\theta_{t-1}, \Sigma_t)$. It is illustrated to be more efficient than uniform kernel and component-wise normal kernel in reflecting the true posterior structure \cite{ref:kernel}. It has been proved to perform well in several dynamic system models \cite{ref:abcsysbio, ref:compare, ref:disease} for the parameter estimation and model selection tasks. The $scaling\in(0,1]$ parameter in \verb|pyABC| will be multiplied to the covariance to produce a `thinner-distributed' perturbation result.

@@

The kernel experiment is designed to explore the efficiency of SMC on our dynamical systems. Given the same fixed threshold schedule, kernels that need less total samples and have higher acceptance rates would be our preference. The experiments compared multivariate normal kernel with local MNN kernels with different parameters, using synthetic data to infer back the parameters. The acceptance rate among each time point and total required samples are compared after the experiment to give suggestions on the kernel selection in the real data inference. As the threshold schedule was fixed, the final population should have similar discrepancy to the target data thus here the goodness of fit, i.e. recover of target data trends/features and errors of the inferred parameters compared to true values were not discussed.

@@

To compare kernels, a fixed schedule was used with the final epsilon $\epsilon_t=10.0$. From the total required samples plot (Figure \ref{fig:kernel1}) the efficiency can be compared. Among the tested kernels, the local M-NN with M=50 has the best performance: it requires the least number of samples and
has higher acceptance rates among almost all generations. Local M-NN with M=750 is the slowest kernel.

For local M-NN kernels, generally, a higher M leads to lower acceptance rates and more required particles in each generation. As shown in Figure \ref{fig:acceptance1}, local M-NN with M=750 is the lowest curve. Consequently, if greater M was tested, then it would have even lower acceptance rates; the maximum M=2000 (the whole population is considered) would have the lowest acceptance rates.

@@

    \caption[Total required number of sample for different kernels]%
    {Total required samples of different kernels after 20 populations (2000 particles in each population). Different color represents different generations (bottom to top: population 1 to population 20)}
    \label{fig:kernel1}

@@

    \caption[Acceptance rates of different kernels]%
    {Acceptance rates of different kernels in 20 populations. Each population has 2000 particles}
    \label{fig:acceptance1}

@@

Multivariate normal kernel has a performance between local M-NN M=250 and M=100. It proves that rather than a trivial normal kernel, multivariate normal kernels are more efficient facing the concentrations of joint distributions among multiple parameters \cite{ref:kernel}. `Scaling' option can narrow the distribution calculated from the last population, thus making the kernel more `local' by sampling under a smaller variance. Similar to small M in local M-NN, small `scaling' is also more efficient in our experiment of model 1. Besides the fixed threshold schedule, a median threshold schedule with the same final threshold $\epsilon_{20}=10.0$ was also tested and similar results are observed (Figure \ref{fig:kernel2} and \ref{fig:acceptance2}, Appendix B2).

A more `local' kernel is more likely to give better performance by efficiently sampling around concentrations in parameters' distribution and approximate the posterior distribution. This holds in most cases under a given target final threshold value $\epsilon_t$. However, it may not produce the proper result want: a more local kernel, e.g. local M-NN with small M and multivariate normal with `scaling' is more likely to be stuck in local optimality, as the kernel is less likely to sample particles that are far from local concentrations, although these local optimal are still accepted before even smaller threshold values is demanded (see Section 5.2 for more discussions). One obvious example is presented in Figure \ref{fig:kernel1}, where local M-NN with M=50 suffered from a local optimal: the last generation takes much more samples to meet the required threshold.

@@

To conclude, Local kernels can be fast but unstable in some case. For a general parameter inference of our models, multivariate kernels were preferred, as a good fit is our prior target. Some local kernels are also worth trying after the multivariate normal kernel, e.g. local M-NN with M$\leq 250$ (for a population size of 2000, i.e. 12.5\% of the population) and multivariate normal with scaling, but extra cares must be taken: although they are fast, moving in and moving out of local optimum make them more unstable in efficiency.

@@

where $i$ is the index of data points and $\Delta x_i$ is the discrepancy between  observed data and simulated data at data point $i$, i.e. $\Delta x_i = x_{i, simulated}-x_{i, observed}$. In this case, data points are 12 time points of four variables, i.e. 48 data points in total.

@@

where a weight $w_i$ is assigned to every data point. If data point $i$ has a higher weight, then the data value is regeared to be more informative, i.e. giving more help in inferring the true posteriors. Weights can be either pre-set according to prior knowledge of the problem or updated using an adaptive method. This method is known as adaptive distance.

Adaptive distance is changing the weights of all data points after each generation iteration, trying to assign informative data points higher weights. Additionally, implementation of adaptive in \verb|pyABC| also introduces additional factors $f_i$ multiplied to weights \cite{ref:adpt_dis}

@@

Factor is helpful as an option for efficiency. When some data points are equally informative, the manually pre-defined can make the distance more focused on certain data points, or behave like a normalisation that can balance the scales of different summarised statistics (here are the 48 mean values). There are two appliance case studied: (1) factors used as a normalisation option and (2) factors used to give more focus on some features of the observed data. For (2), as we noticed that the main features (e.g. rapid increase, peaks, fluctuations) are mostly observed in the first half of the time points, i.e. 0 - 30 hpl, factors were tried in later experiments. Such factor scheme assigns data in the first half of the time point with higher factors (75) and data in the second half with lower factors (25).

@@

From Figure \ref{fig:factor}, different types of factor gave different performance. 75:25 factor required much more samples in the first generation (orange bar in the left of Figure \ref{fig:factor}) and also did not improve much in later generations; thus, it is the slowest one. The first half of observed data contained more features and information, assigning it with higher factor would make the algorithm think that it is more important to fit the first half and thus focus on the hard part, and consequently more particles were needed in the inference process. The acceptance rates of 75:25 factor among different generations were also generally slightly lower than the standard. By contrast, range factor required less total samples and gave generally higher acceptance rates than standard, which made it considerably faster. The adaptive distance gave the best efficiency, as it significantly improved the acceptance rates and resulted in even less required samples (right, Figure \ref{fig:factor}).

@@

To conclude, because the factors and weights are directly multiplied to $\Delta x$, and thus the metrics for distance are changed, the direct comparison of the efficiency under the same threshold schedule was unfeasible; we could only compare the results after the same number of generations. Factors gave nearly the same inferred results but required a different number of samples, hence cares must be taken when choosing factor schemes. Adaptive distance did not provide an accurate result as others. As a result, our further implementations used standard distance function, and factors were also tried for possible improvements.

@@

The prior range and distribution can largely affect the inference progress (Figure \ref{fig:size}. (a)). Samples were drawn from a high dimensional parameter space in each sampling process. A wider parameter prior interval in only one dimension can cause the parameter space to be much bigger. Consequently, the exploration and sampling of parameter vectors would take much more efforts. Right plot in Figure \ref{fig:size} also indicates that the convergency is slower for wider prior ranges. As a result, a proper prior interval based on our experience and belief with seeing the data should be set as narrow as possible, as the inference is largely affected by the prior interval and its length. Improper prior could even make the algorithm miss the true posterior \cite{ref:abcsysbio}.

How to specify prior can be tricky from problem to problem. For our task in the parameter inference of regeneration model, biologically motivated priors were used, which were a general reflection of biological constraints. Also, the units of the parameters helped us in identifying the possible scale of the parameters.

@@

Some other hyperparameters, e.g. the number of generations, population size (i.e. the number of particles per generation) were not addressed in this subsection. In the real inference, population size was set to be greater than 2000 to avoid insufficient exploration of the previous population; in some case than high variance or uncertainties in the results were observed, population size as large as 20,000 were used. If the epsilon did not show a convergency trend, the run could be extended to more generations until a stable result was observed.

@@

The results of prior distribution log-uniform [0, 50] is shown in Figure \ref{fig:result123}, which is considered to give a better fit than other prior options. For these models, the acceptance rates and epsilon path are presented in Figure \ref{fig:result123_2}, and the estimated parameters' posterior of model 3 is shown in Figure \ref{fig:para1}. Figure \ref{fig:result123} plots the simulated trajectories of the three models. One thousand particles (each particle is a parameter set) were sampled, and the mean (blue) and 25th to 75th percentile interval (grey) were calculated from the data simulated by these 1000 particles.

@@

Next, a model selection experiment with different prior was conducted, using the same ABC-SMC framework. As in the parameter inference, two distributions (uniform and log-uniform) are tested with three different interval ([0, 25], [0, 50] and [0, 75]). The results show that model 3 wins with 100\% model probability after 30 generations for most runs, except log-uniform [0, 75]. All the model probability results are of great discrepancy (e.g. 100\% model 3 and 0\% for model 1 and 2), so Bayesian factor is not calculated.

Figure \ref{fig:cmp1} shows the model selection process, under log-uniform [0, 50] prior. Model 3 gains total advantage over model 1 and 2 after the seventh population. Together with the conclusions above, model 3 is considered to be the best model to represent the target data so far.

@@

\paragraph{Model 5} Alternatively, we considered promotions to the tnf-$\alpha$ production, i.e. $s_{\alpha\Phi}\Phi$. It is assumed that il-$\beta$ could accelerate the production process. An additional term $f_{\beta\alpha}\beta$ is introduced to production rate, with $f_{\beta\alpha}$ being a new model parameter (positive constant). This model meets the biological context, considering the source of tnf-$\alpha$. The proposed model is written as Equation \ref{eq:model5}.

@@

\begin{figure}[H]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/overfit.png}}
    \end{center}

@@

\begin{figure}[H]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/model345cmp.png}}
    \end{center}

    \caption[Model probabilities of model 3, 4 and 5 in ABC SMC based model comparison]{Model probabilities of model 3, 4 and 5 in ABC SMC based model comparison. Model 5 also wins in further populations (population 25 to 30)}
    \label{fig:model345cmp}
\end{figure}

@@

Based on the model comparison results, we could conclude that model 5 was the best model among the tested models. Uncertainties were observed in early generations, where similar patterns were also observed in the model comparison of model 1, 2 and 3. We considered the drastically changing model probability in early generations is related to the inference ability at high epsilon threshold values (epsilon) and possibly the insufficient exploration of the parameter space. However as we only focused on the best model with small final threshold $\epsilon_T$, the easiest way is to extend the number of generation until epsilon converges, or increase the population size for a more statistically confident probability.

@@

For the best model (model 5), the inferred parameter posterior distribution of model 5 is shown in Figure \ref{fig:model5_para}, the estimated mean values is shown in Table \ref{table:estimated5}. The pair-wise joint density distribution is shown in Figure \ref{fig:model5_para_joint}. Most of the inferred posteriors show a bell-shaped density distribution with a single peak.. A thin and concentrated shape indicates a well-inferred parameter, e.g. $\mu{\Phi}$ and $\kappa_{\Phi\beta}$. Some distributions, however, are skewed to zero, e.g. $\mu_N$ and $s_{\alpha\Phi}$. Concentrations in the posteriors were presented, and using inter-quartiles (shaded interval in Figure \ref{fig:model5_para}) or mean values to summarise the posteriors is feasible in practice uses, e.g. simulating data (Figure \ref{fig:overfit}).

@@

As models 5 won in the model selection, further interest was to explore the parameter-to-model sensitivity of the resultant model. PCA was performed on the last generation of inferred model 5, Figure \ref{fig:pc} shows the total explained variance by each PC. The first PC explained a major part of the variance (91.5\%), and the rest PCs explained less than 5\% of the variance each. PC 5 to PC 12 explained much less variance (less than 0.1\% of the total variance) than the first four PCs. Thus, they were considered to extend much narrower regions on the posterior distributions and could be used to identify the parameters that the model is sensitive to.

@@

Figure \ref{fig:pc_pie} shows the fractions of the PCs explained by individual parameters, and a sum of the fractions in the last 8 PCs, i.e. PC 5 to PC 12. The last PC (PC12) is mostly the linear combination of $a$ (from the equation of neutrophil) and $\mu_\Phi$ (from the equation of macrophage), to which the model is most sensitive. If we conclude the last 8 PCs, it can be found that the biggest portions ($v_{N\Phi}$, $a$, $\kappa_{N\beta}$, $\kappa_{\Phi\beta}$, $\mu_N$, $\mu_\Phi$, $\nu_{N\Phi}$) mostly come from the equation of neutrophil and macrophage i.e. equations of the cells (the un-exploded portion in Figure \ref{fig:pc_pie} a); if we look at the last 8 PCs individually, the same pattern can be observed. This might suggest that the model is more sensitive to cells' dynamic rather than cytokines' dynamics.

>>>>

The inference framework code for the project was developed and tested in a local environment (macOS 10.15.6, see Table \ref{table:local_macine}) at first using a small input, then the functional version was deployed on compute node of ARCHER and Cirrus for parameter inference experiments and model selection tasks. The scaling-up experiment was performed on Cirrus. 
%The profiling and its analysis were performed on a local machine. 
Some necessary development and debug were performed on the remote machines to ensure that experiments run correctly on the compute node.

The ODE model-related code and utilities were first developed to enable reusable functions, variables and data calls. It included the code for the ODE equations, ODE solver, data structures and data format transform functions, etc. By doing this modularisation in separate code files, we could change implementation options and activate ABC SMC easily in another `trigger' file, without creating definition and duplicated code for the models, observed data and parameter priors.

Some missing features were found when using \verb|pyABC|; therefore some modifications were made to the source code of \verb|pyABC|.For example, the laboratory-measured data were not considered `complete' or `tidy' for modelling and analysis purpose: the available time points for cells and cytokines were different, e.g. cell number was not measured at 0.25 hpl and cytokines' expression was not measured at time point 120 hpl. These missing values were treated as `ignored' when calculating the distance between observed data and simulated data, and represented by \verb|NaN|s. However, in the built-in distance calculations, \verb|NaN| was not acceptable. To cope with this, the built-in distance function was modified. Other modified code included some visualisation and result representation enhancements.

Analysis of the program output was also wrapped up into functions which can be easily called after an ABC SMC run is finished, which include multiple visualisations and out of summary statistics of the results. For different tasks, there are also separated analysis code file. Other analysis tools used in the project include built-in database visualisation tool, R code and Microsoft Excel.

@@

    The ABC SMC inference can have widely different performance when using different hyperparameters and implementation options. To firstly test the ability of inference and observe the results of different options, synthetic trajectories with known true parameter values were used as target data. By doing this, we could directly compare the inferred parameters to the true values and compare the simulated trajectories to the true trajectories, then observe the result to see whether the inference is successful and how well the model fits the data. Also, by setting the same final threshold value, the efficiency under different options can be compared and help us to choose proper options when applying ABC SMC.

@@

    \caption[Synthetic data generated with known parameter values]%
    {Synthetic data generated with known parameter values (line plot), compared to experimental data (scatter plot). Known parameter values are obtained from a least square fitting of the observed data. Error bar indicates SEM}
    \label{fig:infer_back_data}

@@

The perturbation kernel is called in the sampling of every particle and determines how the new perturbed particle is chosen; thus it is influential to both computation complexity and the resulting posterior distribution. Generally, a local perturbation kernel may face the risk of being stuck in local modes (e,g., local optima), but it may need less computational operations or could generate a population with a higher acceptance rate. A kernel with a wide variance, or spreading out in a large space could help in resolving the local optima problem by a more thorough exploration of the parameter space. However, it can be more computation-intensive and result in lower acceptance rates. A desired kernel should balance the trade-offs; their properties and criteria were discussed in \cite{ref:kernel}.

There are several common choices of perturbation kernels. Among them, the multivariate normal kernel and local M-nearest neighbour model is preferred to be applied to our models. A covariance matrix $\Sigma_t$ of accepted particles is calculated from the previous generation and used in multivariate normal kernel: $K(\theta|\theta_{t-1})\sim\mathcal{N}(\theta_{t-1}, \Sigma_t)$. It is illustrated to be more efficient than uniform kernel and component-wise normal kernel in reflecting the true posterior structure \cite{ref:kernel}. It has been proved to perform well in several dynamic system models \cite{ref:abcsysbio, ref:compare, ref:disease} for the parameter estimation and model selection tasks. The $scaling\in(0,1]$ option in \verb|pyABC| will be multiplied to the covariance to produce a `thinner-distributed' perturbation result.

@@

The kernel experiment was designed to explore the efficiency of SMC on our dynamical systems. Given the same fixed threshold schedule, kernels that need less number of samples and have higher acceptance rates would be our preference. The experiment compared multivariate normal kernel with local MNN kernels with different options, using synthetic data to infer back the parameters. The acceptance rate among each generation and total required number of sample were compared to give suggestions on the kernels. As the threshold schedule was fixed, the final population should have similar discrepancy to the target data thus here the goodness of fit (i.e. recover of target data trends/features and errors of the inferred parameters compared to true values) were not discussed.

@@

To compare kernels, a fixed schedule of thresholds was used with the final epsilon $\epsilon_t=10.0$. From the total required number of samples plot (Figure \ref{fig:kernel1}) the efficiency can be compared. Among the tested kernels, the local M-NN with M=50 has the best performance: it requires the least number of samples and
has higher acceptance rates among almost all generations. Local M-NN with M=750 is the slowest kernel.

For local M-NN kernels, generally, a higher M leads to lower acceptance rates and more required particles in each generation. As shown in Figure \ref{fig:acceptance1}, local M-NN with M=750 is the lowest curve. Consequently, if a greater M was tested, then it would have even lower acceptance rates; the maximum M=2000 (the whole population is considered) would have the lowest acceptance rates.

@@

    \caption[Total required number of samples for different kernels]%
    {Total required number of samples of different kernels after 20 populations (2000 particles in each population). Different color represents different generations (bottom to top: population 1 to population 20)}
    \label{fig:kernel1}

@@

    \caption[Acceptance rates of different kernels]%
    {Acceptance rates of different kernels in among populations. Each population has 2000 particles}
    \label{fig:acceptance1}

@@

Multivariate normal kernel has a performance between local M-NN M=250 and M=100. It proves that rather than a trivial normal kernel, multivariate normal kernels are more efficient facing the concentrations of joint distributions among multiple parameters \cite{ref:kernel}. `Scaling' option can narrow the distribution calculated from the last population, thus making the kernel more `local' by sampling under a smaller variance. Similar to small M in local M-NN, small `scaling' is also more efficient in our experiment of model 1. Besides the fixed threshold schedule, a median threshold schedule with the same final threshold $\epsilon_{20}=10.0$ was also tested and similar results are observed (Figure \ref{fig:kernel2} and \ref{fig:acceptance2}).

A more `local' kernel is more likely to give better performance by efficiently sampling around concentrations in parameters' distribution and approximate the posterior distribution more quickly. This holds in most cases under a given target final threshold value $\epsilon_t$. However, it may not produce the proper result as  we want: a more local kernel, e.g. local M-NN with small M or multivariate normal with `scaling' is more likely to be stuck in local optima, as the kernel is less likely to sample particles that are far from local concentrations, although these local optima are still acceptable before even smaller threshold values is demanded (see Section 5.2 for more discussions). One obvious example is presented in Figure \ref{fig:kernel1}, where local M-NN with M=50 suffered from a local optimum: the last generation takes much more samples to meet the required threshold.

@@

To conclude, Local kernels can be fast but unstable in some case. For a general parameter inference for our models, multivariate kernels are preferred, as a good fit is our prior target. Some local kernels are also worth trying after the multivariate normal kernel, e.g. local M-NN with M$\leq 250$ (for a population size of 2000, i.e. 12.5\% of the population) and multivariate normal with scaling, but extra cares must be taken: although they are fast, moving in and moving out of local optimum make them unstable in efficiency.

@@

where $i$ is the index of data points and $\Delta x_i$ is the discrepancy between  observed data and simulated data at data point $i$, i.e. $\Delta x_i = x_{i, simulated}-x_{i, observed}$. In this case, data points are 12 time points for four variables, i.e. 48 data points in total.

@@

where a weight $w_i$ is assigned to every data point. If data point $i$ has a higher weight, then the data value is regeared to be more informative, i.e. giving more help in inferring the true posteriors. Weights can be either pre-set according to prior knowledge of the problem or updated using an adaptive method. The later method is known as adaptive distance.

Adaptive distance is changing the weights of all data points after each generation iteration, trying to assign informative data points higher weights. Additionally, implementation of adaptive in \verb|pyABC| also introduces additional factors $f_i$ multiplied to weights \cite{ref:adpt_dis}.

@@

Factor is helpful as an option for efficiency. When some data points are equally informative, the manually pre-defined factor can make the distance more focuse on certain data points, or behave like a normalisation that can balance the scales of different summarised statistics. There are two appliance case studied: (1) factors used as a normalisation option and (2) factors used to give more focus on some features of the observed data. For (2), as we noticed that the main features (e.g. rapid increase, peaks, fluctuations) are mostly observed in the first half of the time points, i.e. 0 - 30 hpl. Such factor scheme assigns data in the first half of the time point with higher factors (75) and data in the second half with lower factors (25).

@@

From Figure \ref{fig:factor}, different types of factor gave different performance. 75:25 factor required much more samples in the first generation (orange bar in the bottom of 25:75 factor plot, left of Figure \ref{fig:factor}) and also did not improve much in later generations; thus, it is the slowest one. The first half of observed data contained more features and information, assigning it with higher factor would make the algorithm think that it is more important to fit the first half and thus focus on the hard part, and consequently more particles were needed in the inference process. The acceptance rates of 75:25 factor among different generations were also generally slightly lower than the standard. By contrast, range factor required less total samples and gave generally higher acceptance rates than standard, which made it considerably faster. The adaptive distance gave the best efficiency, as it significantly improved the acceptance rates and resulted in even less required samples (right, Figure \ref{fig:factor}).

@@

To conclude, because the factors and weights are directly multiplied to $\Delta x$, and thus the metrics for distance are changed, the direct comparison of the efficiency under the same threshold schedule was unfeasible; we could only compare the results after the same number of generations. Factors gave nearly the same inferred results but required different numbers of samples, hence cares must be taken when choosing factor schemes. Adaptive distance did not provide an accurate result as others. As a result, our further implementations used standard distance function, and factors were also tried for possible improvements.

@@

The prior range and distribution can largely affect the inference progress (Figure \ref{fig:size}). Samples were drawn from a high dimensional parameter space in each sampling process. A wider parameter prior interval in only one dimension can cause the parameter space to be much bigger. Consequently, the exploration and sampling of parameter vectors would take much more efforts. Right plot in Figure \ref{fig:size} also indicates that the convergency is slower for wider prior ranges. As a result, a proper prior interval based on our experience and belief without seeing the data should be set as narrow as possible, as the inference is largely affected by the prior interval and its length. Improper prior could even make the algorithm miss the true posterior \cite{ref:abcsysbio}.

How to specify prior can be tricky from problem to problem. For our task in the parameter inference of regeneration model, biologically motivated priors were used, which were a general reflection of biological constraints. Also, the units of the parameters helped us in identifying the possible scales of the parameters.

@@

Some other hyperparameters, e.g. the number of generations, population size (i.e. the number of particles per generation) were not addressed in this subsection. In the real inference, population size was set to be greater than 2000 to avoid insufficient exploration of the previous population; in some case than high variance or uncertainties in the results were observed, as large as 20,000 were used as population size. If the epsilon did not show a convergency trend, the run could be extended to more generations until a stable result was observed.

@@

The results of prior distribution log-uniform [0, 50] is shown in Figure \ref{fig:result123}, which is considered to give a better fit than other prior options. For these models, the acceptance rates and epsilon path are presented in Figure \ref{fig:result123_2}, and the estimated parameters' posterior of model 3 is shown in Figure \ref{fig:para1}. Figure \ref{fig:result123} plots the simulated trajectories of the three models. One thousand particles (each particle is a parameter set) were sampled for the last generation, and the mean (blue) and 25th to 75th percentile interval (grey) were calculated from the data simulated by these 1000 particles.

@@

Next, a model selection experiment with different prior was conducted, using the same ABC-SMC framework. As in the parameter inference, two distributions (uniform and log-uniform) are tested with three different interval ([0, 25], [0, 50] and [0, 75]). The results show that model 3 wins with 100\% model probability after 30 generations for most runs, except log-uniform [0, 75]. All the model probability results are of great discrepancy (e.g. 100\% for model 3 and 0\% for model 1 and 2), so Bayesian factor is not calculated.

Figure \ref{fig:cmp1} shows the model selection process, under log-uniform [0, 50] prior. Model 3 gains total advantage over model 1 and 2 after the seventh population. Together with the conclusions above, model 3 is considered to be the best model to represent the target data so far. Considering the motivations in proposing model 2 and 3, the result shows that a decaying $\lambda$ term is reasonable in both theory and practice; simplification of $i_{\beta\Phi}$ term gives an even better model (i.e. model 3), which means that the inhibition effect from macrophage can indeed be ignored.

@@

\paragraph{Model 5} Alternatively, we considered promotions to the tnf-$\alpha$ production, i.e. $s_{\alpha\Phi}\Phi$. It is assumed that il-$\beta$ could accelerate the production process. An additional term $f_{\beta\alpha}\beta$ is introduced to production rate, with $f_{\beta\alpha}$ being a new model parameter (positive constant). This model meets the biological context considering the source of tnf-$\alpha$. The proposed model is written as Equation \ref{eq:model5}.

@@

\begin{figure}[ht!]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/overfit.png}}
    \end{center}

@@

\begin{figure}[ht!]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/model345cmp.png}}
    \end{center}

    \caption[Model probabilities of model 3, 4 and 5]{Model probabilities of model 3, 4 and 5 in ABC SMC based model comparison. Model 5 also wins in further populations (population 25 to 30)}
    \label{fig:model345cmp}
\end{figure}

@@

Based on the model comparison results, we could conclude that model 5 was the best model among the tested models. 
It proves that the extra term $f_{\beta\alpha}$ not only makes sense regarding the source of tnf-$\alpha$, but also gives a better model than phenomenological term $d_{\beta\alpha}\beta$.
Uncertainties were observed in early generations, where similar patterns were also observed in the model comparison of model 1, 2 and 3. We considered the drastically changing model probability in early generations is related to the inference ability at high epsilon threshold values (epsilon) and possibly the insufficient exploration of the parameter space. However as we only focused on the best model with small final threshold $\epsilon_T$, the easiest way is to extend the number of generation until epsilon converges, or increase the population size for a more statistically confident probability.

@@

For the best model (model 5), the inferred parameter posterior distribution of model 5 is shown in Figure \ref{fig:model5_para}, the estimated mean values is shown in Table \ref{table:estimated5}. The pair-wise joint density distribution is shown in Figure \ref{fig:model5_para_joint}. Most of the inferred posteriors show a bell-shaped density distribution with a single peak.. A thin and concentrated shape indicates a well-inferred parameter, e.g. $\mu_{\Phi}$ and $\kappa_{\Phi\beta}$. Some distributions, however, are skewed to zero, e.g. $\mu_N$ and $s_{\alpha\Phi}$. Concentrations in all the posteriors were presented, and using inter-quartiles (shaded interval in Figure \ref{fig:model5_para}) or mean values to summarise the posteriors is feasible in practice uses, e.g. simulating data (Figure \ref{fig:overfit}).

@@

As models 5 won in the model selection, further interest was to explore the parameter-to-model sensitivity of the resultant model. PCA was performed on the last generation of the inferred model 5, Figure \ref{fig:pc} shows the total explained variance by each PC. The first PC explained a major part of the variance (91.5\%), and the rest PCs explained less than 5\% of the variance each. PC 5 to PC 12 explained much less variance (less than 0.1\% of the total variance) than the first four PCs. Thus, they were considered to extend much narrower regions on the posterior distributions and could be used to identify the parameters that the model is most sensitive to.

@@

Figure \ref{fig:pc_pie} shows the fractions of the PCs explained by individual parameters, and a sum of the fractions in the last 8 PCs, i.e. PC 5 to PC 12. The last PC (PC12) is mostly the linear combination of $a$ (from the equation of neutrophil) and $\mu_\Phi$ (from the equation of macrophage), to which the model is most sensitive. If we conclude the last 8 PCs, it can be found that the biggest portions ($v_{N\Phi}$, $a$, $\kappa_{N\beta}$, $\kappa_{\Phi\beta}$, $\mu_N$, $\mu_\Phi$, $\nu_{N\Phi}$) mostly come from the equation of neutrophil and macrophage i.e. equations of the cells (the un-exploded wedges in Figure \ref{fig:pc_pie} a); if we look at the last 8 PCs individually, the similar pattern can be observed. This might suggest that the model is more sensitive to cells' dynamic rather than cytokines' dynamics.

### 2020-08-20 introduction/introduction.tex
Proofread

Zebrafish can regenerate its spinal cord after injury; tissue regeneration observed at the injury site is a complex biological process with interacting parts and dynamical feedback between different types of cells and molecules. These cells and molecules are involved in the inflammation and repair dynamically, and a complex interacting network can be drawn according to it.

@@

Mathematical models and techniques can help us to resolve the complexity in this dynamic system; however, how to derive the parameter estimates of the models remains a challenging work. In this dissertation, our interest lies in the modelling of the regeneration process that happens around the lesion site, especially in the parameter estimates and comparison of possible models. According to available researches \cite{ref:Tsarouchas}, a mathematical model was first proposed, but the addressed interactions remain to be precisely defined, and quantitatively examined in experiments. We want to find the best parameter sets of the model using the published data, and ranking models under specific metrics and possibly find the best-fit model. In these steps, models are expected to retain mathematical and biological significance regarding the complex interactions, and differential equations are suitable for this.

@@

For the parameter estimation tasks, optimisation-based methods usually have trouble in determining global optima from local optimal, and noise in the observed data may lead to over-fitting. Compared to optimisation-based methods, Bayesian inference techniques can help to relieve these two concerns\cite{ref:abcsysbio}. Moreover, approximate inference methods (Approximate Bayesian Computation, ABC) make likelihood-free inference possible, which is suitable for our task as writing down likelihood as a function expression for the dynamic system can be hard.

Such methods introduce computation-intensive sampling-and-test patterns, where a massive number of samples are drawn and followed by complex computations. As the sampling is independent of each other in a certain period, the parallel implementation of the algorithm becomes feasible and is expected to accelerate the parameter inference on a large scale when using HPC resources.

This dissertation mainly focuses on the implementations and experiments of parameter estimation using ABC SMC (sequential Monte-Carlo) method, and discussions on the results. Additionally, performance experiments illustrated the strong-scaling performance of the algorithm. Chapter 2 talks about the background information and theories, Chapter 3 shows how the mathematical models are derived from hypothesis, and Chapter 4 describes the ABC SMC implementations and experiments on the models in detail. A brief scaling-up performance experiment is included in Chapter 5, and Chapter 6 summarises the conclusions and states the future works on this topic.

>>>>

Zebrafish can regenerate its spinal cord after injury; tissue regeneration observed at the injury site is a complex biological process with many interacting parts and dynamical feedback between different types of cells and molecules. These cells and molecules are involved in the inflammation and repair dynamically, and a complex interacting network can be drawn according to it.

@@

Mathematical models and techniques can help us to resolve the complexity in this dynamic system; however, how to derive the parameter estimates of the models remains a challenging work. In this dissertation, our interest lies in the modelling of the regeneration process that happens around the lesion site, especially in the parameter estimates and comparison of possible models. According to available researches \cite{ref:Tsarouchas}, a mathematical model was first proposed, but the addressed interactions remain to be precisely defined, and quantitatively examined in experiments. We want to explore the uncertainties in parameter estimates using the published data, ranking models under specific metrics and possibly find the best-fit model. In these steps, models are expected to retain mathematical and biological significance regarding the complex interactions, where differential equations are suitable for this.

@@

For the parameter estimation task, optimisation-based methods usually have trouble in determining the global optima, and noise in the observed data may lead to over-fitting. Compared to optimisation-based methods, Bayesian inference techniques can help to relieve these two concerns \cite{ref:abcsysbio}. Moreover, approximate inference methods (Approximate Bayesian Computation, ABC) make likelihood-free inference possible, which is suitable for our dynamic system as writing down likelihood as a function expression can be hard.

Such methods introduce computation-intensive sampling-and-test patterns, where a massive number of samples are drawn and followed by complex computations. As the sampling is independent of each other in a certain period, the parallel implementation of the algorithm becomes feasible. It is expected to accelerate the parameter inference on a large scale when using HPC resources.

This dissertation mainly focuses on the implementations and experiments of parameter estimation using ABC SMC (sequential Monte-Carlo) method, and discussions on the results. Additionally, performance experiments illustrates the strong-scaling performance of the algorithm. Chapter 2 talks about the background information and theories, Chapter 3 shows how the mathematical models were derived from hypotheses, and Chapter 4 describes the ABC SMC implementations and experiments in detail. A brief scaling-up performance experiment is included in Chapter 5, and Chapter 6 summarises the conclusions and states the future works regarding this topic.
